---
title: "IMUL restraint analysis_workflow critique"
author: "Hanzhi Wang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = c(CRAN = "https://cran.r-project.org"))
```

# ReadMe
Input include two csv files: "lena-imu-compiled.csv" and "unrestrained_data_211_date_converted.csv"

Output include two tables (saved as docx.) and three figures (saved as .eps)

# Preliminaries 

Install packages if they haven't installed
```{r, message=F}
if (!("DescTools" %in% installed.packages())) install.packages("DescTools")
if (!("psych" %in% installed.packages())) install.packages("psych")
if (!("tidyverse" %in% installed.packages())) install.packages("tidyverse")
if (!("lme4" %in% installed.packages())) install.packages("lme4")
if (!("performance" %in% installed.packages())) install.packages("performance")
if (!("lmerTest" %in% installed.packages())) install.packages("lmerTest")
if (!("Cairo" %in% installed.packages())) install.packages("Cairo")
if (!("flextable" %in% installed.packages())) install.packages("flextable")
if (!("officer" %in% installed.packages())) install.packages("officer")
if (!("report" %in% installed.packages())) install.packages("report")
```

Load packages 
```{r}
library(DescTools)
library(psych)
library(tidyverse)
library(lme4)
library(performance)
library(lmerTest)
library(report)
```

set path
```{r set path}
setwd("/Users/hanzhiwang/Library/CloudStorage/GoogleDrive-hwang567@ucr.edu/My Drive/Cousework/2024 Fall/PSYC 211_Stats/Project")

```

Load original data
```{r import data}
df_lena0 <- 
  read.csv("lena-imu-compiled.csv")

df_restraint0 <- 
  read.csv("unrestrained_data_211_date_converted.csv")

```

Check data type
```{r check data type, eval=FALSE}
glimpse(df_lena0)
glimpse(df_restraint0)
```

Create a new session-level data frame "df_comb", and copy unique_session id into rows
```{r create a session-level dataframe with subject id and session}
df_comb <- df_restraint0 %>%
  distinct(unique_id, .keep_all = TRUE) %>%  
  select(unique_id, id, session,agemo,sitter,crawler,walker,age_group)   
```

Import sex, race, and ethnicity data from lena dataset into a new data frame
```{r get sex, race, ethnicity data from lena dataset}
df_lena_demo <- df_lena0 %>%
  distinct(unique_id, .keep_all = TRUE) %>%  
  select(id, session, infant_sex, infant_race, infant_ethnicity)
```

Import demographic data into restraint data frame
```{r import demographic data into restraint data frame}
df_comb <- left_join(df_comb,df_lena_demo,by=c("id","session"))
```

Color preset for different locomotor groups
```{r}
loc_colors <- c(
    "tomato",
    "chartreuse4",
    "cyan3",
    "mediumpurple1")
color_noncrw <- "red"
color_nonmob <- "red"
color_mob <- "navy"

```

# Data clean
Noticed in ethnicity some are coded as "Hispanic or Latino", while others as "Hispanic/Latino". Make it consistent
```{r change from "Hispanic or Latino" to "Hispanic/Latino"}
df_comb$infant_ethnicity <- ifelse(df_comb$infant_ethnicity=="Hispanic or Latino","Hispanic/Latino",
                                   no = df_comb$infant_ethnicity)
```

Exclude participant 104 from later analysis because this was a pilot session
```{r}
df_comb <- df_comb %>%
  filter(id != 104)
```

Participant 180's sex is missing. This is a boy
```{r}
df_comb[df_comb$id==180,"infant_sex"] <- "male"
```


# Basic information of each session
Count how many subjects
```{r how many subjects}
num_subj <- n_distinct(df_comb$id)
num_subj
# 66 subjects, 143 sessions
```

Count sex, race, ethnicity
```{r sex, race, ethnicity}
# get a data frame with each row being a participant (instead of session)
df_subj <- df_comb %>%
  distinct(id, .keep_all = TRUE)

table(df_subj$infant_sex,useNA = "ifany")
df_subj[is.na(df_subj$infant_sex),]

table(df_subj$infant_race,useNA = "ifany")
df_subj[is.na(df_subj$infant_race),]

table(df_subj$infant_ethnicity,useNA = "ifany")
df_subj[is.na(df_subj$infant_ethnicity),]


```

How many sessions in younger/older group
```{r how many sessions in younger/older group}
table(df_comb$age_group)
# 67 in younger, and 76 in older group

```

How many subjects in younger/older group
```{r how many subjects in younger/older group}
table(df_subj$age_group)
# 30 in younger, and 36 in older group

```

Age range
```{r}
range(df_comb$agemo[df_comb$age_group=="Younger"])
mean(df_comb$agemo[df_comb$age_group=="Younger"])
sd(df_comb$agemo[df_comb$age_group=="Younger"])

range(df_comb$agemo[df_comb$age_group=="Older"])
mean(df_comb$agemo[df_comb$age_group=="Older"])
sd(df_comb$agemo[df_comb$age_group=="Older"])

```


Independence test of age group and sex/race/ethnicity
```{r}
table(df_subj$age_group, df_subj$infant_sex)
table(df_subj$age_group, df_subj$infant_race)
table(df_subj$age_group, df_subj$infant_ethnicity)
chisq.test(df_subj$age_group, df_subj$infant_sex)
chisq.test(df_subj$age_group, df_subj$infant_race)
chisq.test(df_subj$age_group, df_subj$infant_ethnicity)

```

Make the agemo into integers (e.g., 4, 5 mos), and count how many sessions in 4/5/.../14 age month group
```{r how many sessions in 4/5/.../14 age month group}
df_comb$age_mo_grp <- ifelse(df_comb$agemo>3.5 & df_comb$agemo<4.5,4,
                            ifelse(df_comb$agemo>4.5 & df_comb$agemo<5.5,5,
                                   ifelse(df_comb$agemo>5.5 & df_comb$agemo<6.5,6,
                                          ifelse(df_comb$agemo>6.5 & df_comb$agemo<7.5,7,
                                                 ifelse(df_comb$agemo>10.5 & df_comb$agemo<11.5,11,
                                                        ifelse(df_comb$agemo>11.5 & df_comb$agemo<12.5,12,
                                                               ifelse(df_comb$agemo>12.5 & df_comb$agemo<13.5,13,
                                                                      ifelse(df_comb$agemo>13.5 & df_comb$agemo<14.5,14,NA))))))))


age_mo_dist <- table(df_comb$age_mo_grp)
cat("Table: Session numbers in different age month groups")
age_mo_dist

```


Categorize sessions into walker/crawler/sitter for future analysis
```{r catergorize sessions into walker/crawler/sitter}
# Note, for non-sitters but crawlers, I treat them as crawlers.

df_comb$loc_cat <- ifelse(df_comb$walker=="Walker"& !is.na(df_comb$walker),"walker",
                          ifelse(df_comb$crawler=="Crawler"& !is.na(df_comb$crawler),"crawler",
                                 ifelse(df_comb$sitter=="Sitter"& !is.na(df_comb$sitter),"sitter",
                                  ifelse(df_comb$sitter=="Non-Sitter","non-sitter",NA))))

table(df_comb$loc_cat)
table (df_comb$sitter,df_comb$age_mo_grp, useNA = "ifany")
table (df_comb$crawler,df_comb$age_mo_grp, useNA = "ifany")
table (df_comb$walker,df_comb$age_mo_grp, useNA = "ifany")
```


# Dataset distribution figure
Give temporal id to each subject, for graphing purpose
```{r}
df_comb <- df_comb %>%
  mutate(temp_id = dense_rank(id))

```

Divide the dataset into four locomotor status groups and draw a graph showing dataset distribution
```{r visualize repeated measures with four locomotor status groups, fig.width=7, fig.height=10}

ids <- unique(df_comb$id)

df_comb$loc_cat <- factor(df_comb$loc_cat, levels = c("non-sitter", "sitter", "crawler","walker"))

p_dataset_4Loc <- ggplot(df_comb, aes(x=age_mo_grp, y=temp_id,color = factor(loc_cat))) +
  geom_line(aes(group = temp_id), color = "black") +
  geom_point(shape=16, size=3, alpha=1)+
  scale_color_manual(values=loc_colors) +
  scale_x_continuous(breaks = seq(4,14,by = 1))+
  scale_y_continuous(breaks = c(seq(1, 66, by = 1)),
                     labels = ids) +
  labs(x="Age group (months)",
       y="Subject id",
       title = "Figrue 2 Dataset distribution",
       color = "Locomotor status")+
  theme(panel.grid = element_blank(),
        panel.background = element_blank(),
        text=element_text(size=10),
        axis.text.x=element_text(size=9,color="black"),
        axis.text.y=element_text(size=9,color="black"))
print(p_dataset_4Loc)

ggsave("Figure 2_Dataset_Four Loc.eps", width = 10, height = 25, units = "cm")
```


# Calculate restrained time proportion
calculate total time duration of each session
```{r calcualte total time duration of each session}
# total time should be the row number of each session in df_restraint0 *15/60
df_comb$time_ttl <- (map2_int(df_comb$id, df_comb$session, 
                             ~ sum(df_restraint0$id == .x & df_restraint0$session == .y))*15/60)

```

calculate nap/excluded time period of each session
```{r calculate nap/excluded time period of each session}
df_comb$time_na <- (map2_int(df_comb$id, df_comb$session, 
                             ~ sum(df_restraint0$id == .x & df_restraint0$session == .y & 
                                   (df_restraint0$nap_period==1 | df_restraint0$exclude_period==1)))*15/60)


```

calculate useful time, which is total time subtracts useless time
```{r calculate useful time, which is total time subtract useless time}
df_comb$time_use <- df_comb$time_ttl - df_comb$time_na

```

calculate restrained time proportion among the useful time
```{r calculate restraint time among the useful time}
df_comb$time_rstr <- (map2_int(df_comb$id, df_comb$session, 
                             ~ sum(df_restraint0$id == .x & df_restraint0$session == .y & 
                                   df_restraint0$nap_period==0 & df_restraint0$exclude_period==0 &
                                   df_restraint0$unrestrained=="No"))*15/60)
df_comb$prop_rstr <- df_comb$time_rstr/df_comb$time_use

```

Time durations descriptive results
```{r}
range(df_comb$time_ttl)
mean(df_comb$time_ttl)
sd(df_comb$time_ttl)

range(df_comb$time_na)
mean(df_comb$time_na)
sd(df_comb$time_na)

range(df_comb$time_use)
mean(df_comb$time_use)
sd(df_comb$time_use)

```

# Create age_base, age_change variables and save to df_comb
create age_baseline variable
```{r}
df_comb <- df_comb %>%
    group_by(id) %>%
    mutate(age_base = first(agemo)) %>%
    ungroup()
```

create age_change variable
```{r}
df_comb$age_change <- df_comb$agemo-df_comb$age_base
```


# Description of restrained time proportion
## Scatter plot
```{r}
p_jitter_all <- ggplot(df_comb, aes(x = age_mo_grp, y = prop_rstr)) +
  geom_jitter(width = 0.2, height = 0, color = "blue") +
  geom_smooth(method = "lm",se=T,color = "blue",fill = "blue", alpha = 0.5)+
  scale_x_continuous(breaks = c(seq(4,14, by = 1))) +
  scale_y_continuous(breaks = c(seq(0, 1, by = 0.2)),limits = c(0,1)) +
  labs(title = "Figure 3 Age-related change in RTP (with outliers)", 
       x = "Age Group (months)", 
       y = "Restrained Time Proportion (%)")+ 
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        text=element_text(size=10),
        axis.text.x=element_text(size=9,color="black"),
        axis.text.y=element_text(size=9,color="black"))
  
print(p_jitter_all)
```

# Exclude outliers from both groups (not in each age_month group) based on restrained time proportion
```{r}
outliers_list_older <- 
  DescTools::Outlier(df_comb$prop_rstr[df_comb$age_group=="Older"], method = "boxplot")
outliers_list_younger <- 
  DescTools::Outlier(df_comb$prop_rstr[df_comb$age_group=="Younger"], method = "boxplot")

df_comb_excl <- 
  df_comb[!(df_comb$prop_rstr %in% outliers_list_older | df_comb$prop_rstr %in% outliers_list_younger),]

# change the locomotor categorization into mobile and non-mobile for future use
df_comb_excl_mob <- df_comb_excl %>% 
  mutate(loc_cat = ifelse(loc_cat=="non-sitter"|loc_cat=="sitter","nonmobile","mobile"))
```
Excluded a non-sitter session (178_03) and walker (181_03).

# Descriptive results of RTP (restrained time proportion), after data exclusion
RTP among all sessions
```{r}
describe(df_comb_excl$prop_rstr)
```

RTP in different age month groups (e.g., 4-,5-month-old etc)
```{r}
dscrpt_age <- df_comb_excl %>%
  group_by(age_mo_grp) %>%
  summarize(
    mean = mean(prop_rstr, na.rm=TRUE),
    median = median(prop_rstr, na.rm=TRUE),
    sd = sd(prop_rstr, na.rm=TRUE),
    skew = skew(prop_rstr, na.rm = TRUE),
    kurtosi = kurtosi(prop_rstr, na.rm = TRUE),
    n = n()
  )
print (dscrpt_age)

table_age <- dscrpt_age %>%
  select(age_mo_grp,mean, sd,n) %>%
  as.data.frame() %>%
  round(3)

desc_table <- 
  flextable::flextable(table_age) %>% 
  flextable::set_header_labels(
    age_mo_grp = "Age Group",
    mean = "Mean",
    sd = "SD",
    n = "N") %>%
  flextable::autofit()

doc <- 
  officer::read_docx() %>%
  flextable::body_add_flextable(desc_table) %>%
  print(target = "Table_Age M SD.docx") 
# print the table into a word document with APA style

```

RTP in different age groups (i.e., younger, older)
```{r}
dscrpt_ageGrp <- df_comb_excl %>%
  group_by(age_group) %>%
  summarize(
    mean = mean(prop_rstr, na.rm=TRUE),
    median = median(prop_rstr, na.rm=TRUE),
    sd = sd(prop_rstr, na.rm=TRUE),
    skew = skew(prop_rstr, na.rm = TRUE),
    kurtosi = kurtosi(prop_rstr, na.rm = TRUE)
  )
print (dscrpt_ageGrp)
```

RTP in different locomotor groups (non-sit, sit, crawl, walk)
```{r}
dscrpt_loc <- df_comb_excl %>%
  group_by(loc_cat) %>%
  summarize(
    mean = mean(prop_rstr, na.rm=TRUE),
    median = median(prop_rstr, na.rm=TRUE),
    sd = sd(prop_rstr, na.rm=TRUE),
    skew = skew(prop_rstr, na.rm = TRUE),
    kurtosi = kurtosi(prop_rstr, na.rm = TRUE)
  )
print (dscrpt_loc)
```

RTP in different mobility groups (mobile, non-mobile)
```{r}
dscrpt_mob <- df_comb_excl_mob %>%
  group_by(loc_cat) %>%
  summarize(
    mean = mean(prop_rstr, na.rm=TRUE),
    median = median(prop_rstr, na.rm=TRUE),
    sd = sd(prop_rstr, na.rm=TRUE),
    skew = skew(prop_rstr, na.rm = TRUE),
    kurtosi = kurtosi(prop_rstr, na.rm = TRUE),
    n = n()
  )
print (dscrpt_mob)

table_mob <- dscrpt_mob %>%
  select(loc_cat,mean, sd,n) %>%
  as.data.frame()
table_mob$mean <- round(table_mob$mean,3)
table_mob$sd <- round(table_mob$sd,3)

desc_table <- 
  flextable::flextable(table_mob) %>% 
  flextable::set_header_labels(
    loc_cat = "Mobility",
    mean = "Mean",
    sd = "SD",
    n = "N") %>%
  flextable::autofit()

doc <- 
  officer::read_docx() %>%
  flextable::body_add_flextable(desc_table) %>%
  print(target = "Table_Mobility M SD.docx")
```


# Multilevel: age-related change
~1+age_baseline+(1|participant)
~1+age_baseline+age_change+(1|participant)
~1+age_baseline+age_change+(1+age_change|participant)

## Modeling
```{r}
mod.age1 <- lmer(prop_rstr ~ 1 + age_base + (1|id), data = df_comb_excl)
summary(mod.age1)

mod.age2 <- lmer(prop_rstr ~ 1 + age_base + age_change + (1|id), data = df_comb_excl)
summary(mod.age2)

mod.age3 <- lmer(prop_rstr ~ 1 + age_base + age_change + (1+age_change|id), data = df_comb_excl)
summary(mod.age3)

# Check assumptions
# check_collinearity(mod.age1)
# check_collinearity(mod.age2)
# check_collinearity(mod.age3)
# diagnostic_plots <- plot(check_model(mod.age1, panel = FALSE))
# diagnostic_plots <- plot(check_model(mod.age2, panel = FALSE))
# diagnostic_plots <- plot(check_model(mod.age3, panel = FALSE))
# diagnostic_plots[[1]]
# diagnostic_plots[[2]]
# diagnostic_plots[[3]]
# diagnostic_plots[[4]]
# diagnostic_plots[[5]]
# diagnostic_plots[[6]]

report(mod.age1)
cat('\n')
report(mod.age2)
cat('\n')
report(mod.age3)
```

Compare models
```{r}
anova(mod.age1,mod.age2)
anova(mod.age1,mod.age3)
anova(mod.age2,mod.age3)
```

## Scatter plot
```{r}
p_jitter_all <- ggplot(df_comb_excl, aes(x = age_mo_grp, y = prop_rstr)) +
  geom_jitter(width = 0.2, height = 0, color = "blue") +
  #geom_smooth(method = "lm",se = T,color = "blue",fill = "blue", alpha = 0.5)+
  stat_summary(fun = mean, geom = "crossbar", width = 0.8)+
  scale_x_continuous(breaks = c(seq(4,14, by = 1))) +
  scale_y_continuous(breaks = c(seq(0, 1, by = 0.2)),limits = c(0,1)) +
  labs(title = "Figure 3 Age-related change in RTP", 
       x = "Age Group (months)", 
       y = "Restrained Time Proportion (%)")+ 
  theme_minimal()+
  theme(panel.grid.minor = element_blank(),
        text=element_text(size=10),
        axis.text.x=element_text(size=9,color="black"),
        axis.text.y=element_text(size=9,color="black"))
  
print(p_jitter_all)

ggsave("Figure 3_Age change.eps", 
       # device = Cairo::cairo_ps, 
       width = 14, height = 8, units = "cm")
```

## Age-related change in younger group only, same models as above
```{r}
df_comb_excl_younger <- df_comb_excl %>%
  subset(age_group == "Younger")

mod.age_younger1 <- lmer(prop_rstr ~ 1 + age_base + (1|id), data = df_comb_excl_younger)
mod.age_younger2 <- lmer(prop_rstr ~ 1 + age_base + age_change+ (1|id), data = df_comb_excl_younger)
mod.age_younger3 <- lmer(prop_rstr ~ 1 + age_base + age_change+ (1+age_change|id), data = df_comb_excl_younger)
summary(mod.age_younger1)
summary(mod.age_younger2)
summary(mod.age_younger3)

anova(mod.age_younger1,mod.age_younger2)
anova(mod.age_younger1,mod.age_younger3)
anova(mod.age_younger2,mod.age_younger3)

report(mod.age_younger2)

```

## Age-related change in older group only, same models as above
```{r}
df_comb_excl_older <- df_comb_excl %>%
  subset(age_group == "Older")

mod.age_older1 <- lmer(prop_rstr ~ 1 + age_base + (1|id), data = df_comb_excl_older)
mod.age_older2 <- lmer(prop_rstr ~ 1 + age_base + age_change+ (1|id), data = df_comb_excl_older)
mod.age_older3 <- lmer(prop_rstr ~ 1 + age_base + age_change+ (1+age_change|id), data = df_comb_excl_older)
summary(mod.age_older1)
summary(mod.age_older2)
summary(mod.age_older3)

anova(mod.age_older1,mod.age_older2)
anova(mod.age_older1,mod.age_older3)
anova(mod.age_older2,mod.age_older3)

report(mod.age_older2)

```


# Multilevel: mobility-related change

## Modeling
```{r}
mod.mob1 <-lmer(prop_rstr ~ 1 + age_base + age_change + loc_cat+ (1|id), data = df_comb_excl_mob)
summary(mod.mob1)
mod.mob2 <-lmer(prop_rstr ~ 1 + age_base + age_change + loc_cat+ (1+age_change|id), data = df_comb_excl_mob)
summary(mod.mob2)
mod.mob3 <-lmer(prop_rstr ~ 1 + age_base + age_change + loc_cat+ (1+loc_cat|id), data = df_comb_excl_mob)
summary(mod.mob3)

# Check assumptions
# diagnostic_plots <- plot(check_model(mod.mob1, panel = F))
# check_collinearity(mod.mob1) 
# diagnostic_plots <- plot(check_model(mod.mob2, panel = F))
# check_collinearity(mod.mob2)
# check_collinearity(mod.mob3)
# diagnostic_plots[[1]]
# diagnostic_plots[[2]]
# diagnostic_plots[[3]]
# diagnostic_plots[[4]]
# diagnostic_plots[[5]]
# diagnostic_plots[[6]]

anova(mod.mob1,mod.mob2)
anova(mod.mob1,mod.mob3)
anova(mod.mob2,mod.mob3)

report(mod.mob1)
```

## Scatter plot
```{r}
p_jitter_mobile <-ggplot(df_comb_excl_mob, aes(x = age_mo_grp, y = prop_rstr, color = loc_cat)) +
  geom_jitter(width = 0.2, height = 0) +
  geom_smooth(method = "lm", se = F) +
  # scale_linetype_manual(values = c("crawler" = "solid", "non-crawler" = "dashed")) +
  scale_color_manual(values = c("mobile" = color_mob, "nonmobile" = color_nonmob)) +
  # scale_shape_manual(values = c("crawler" = 16, "non-crawler" = 1)) + 
  scale_x_continuous(breaks = c(seq(4,14, by = 1))) +
  scale_y_continuous(breaks = c(seq(0, 1, by = 0.2)),limits = c(0,1)) +
  labs(title ="Figure 4.4 nonmobile vs. mobile in RTP",
       x = "Age Group (months)",
       y = "Restrained Time Proportion (%)") +
  theme_minimal()+
  theme(panel.grid.minor = element_blank())
print (p_jitter_mobile)

ggsave("Figure 4.4_nonmobile vs mobile in RTP.eps", 
       # device = Cairo::cairo_ps, 
       width = 14, height = 8, units = "cm")
```

